\section{Sensor data visualization}


\subsection{What is sensor data visualization ?}
% Data visualization through dashboards (def)

Stephen Few define a dashboard as "a visual display of the most
important information needed to achieve one or more objectives;
consolidated and arranged on a single screen so the information can be
monitored at a glance"\cite{few:dashboard} In this paper, we will
discuss about multiple dashboards, i.e. succession of screens, in
order to include in our study the sensor data rendering organised in
separate unit of visualization.

% What kind of visualizations ?
According to this definition, the visualizations used for this purpose
should be a graphical representation of the data to treat. Apart from
the well-known generic charts (e.g. line chart, bar chart, pie chart,
...), a huge amount of specialized visualizations is available for
specific data and/or task.
% Static vs dynamic
Such graphical widgets may present a static representation of the
data, as an assessment of a situation, or may offer one to interact
with it with an aim of exploration, to search a more relevant data for
each use, or to offer both an overview and a way to reduce the scope
of the visualization.
% Nb of dimensions
Besides the ornamental criteria which only consist on a effect to
change the aspect of the widget without impacting it capabilities
(eg. 3D pie chart), the number of dimensions displayed play a part in
the final use of the visualizations.  Indeed, there may be one
dimension (e.g. a simple display of your current speed on your car
dashboard), two dimensions (as the generic charts above), or more
(e.g. a country map with a color gradient to weight the population).
Among the dimensions of a given visualization, one or several has to
be used to categorize the input data. The other dimensions are the
values to display.  For example, a line chart represents values
according to the time where a pie chart display, for each entry of a
given categorization, the volume or percentage of matching data. The
previously mentioned three dimension map displays a value, the amount
of human being, in function of a complex category : the combination of
a latitude and a longitude.  In addition, due to ergonomic
constraints, a visualization may impose a limit on (i) the number of
different category displayable, e.g. a pie chart should have eight
portions maximum; (ii) the number of value dimensions, e.g. a line
chart should not represent more than ten lines to be readable (iii) on
both category and value number, e.g. a bar chart should displays
maximum five values per category, and the number of category should be
limited according to the screen size so that the user can still
distinguish each bar from the others.

% For what kind of data ?
Raw data produce by physical sensors are essentially timed values. The
type of the value (eg. string, numerical, ...) and the unit of
measurement in which it is expressed have an influence on the suitable
possibilities of visualization.  For example, given a window status
sensor emitting either "opened" or "close" every minute, the use of a
pie chart would not be relevant. On the other hand, knowing that the
values are temperature measure may lead to a continuous representation
(e.g. line chart) instead of a discrete one (e.g. a bar chart), or
even to a specialized widget for this kind of data.
% Virtual sensor & Data composition
Data can also be the result of a calculation, which can be as simple
as the transformation of an electrical tension to a temperature, or a
more complex one such as a composition of several dataset to infer a
new one.

% For what purposes ?
The intention pursued in the dashboard construction is directly linked
to design choices and to the selection visualization among the amount
of possibilities available.  Indeed, two different widgets, fed with
the exact same data, may highlight specific aspect of it and lead to
different domains of interpretation. For example, the use of a line
chart implicitly infer the impression of continuity and so leads to
focus the interpretation on the variations in the values while, in a
other hand, using a bar chart will result in an analysis of the
comparison between several values for the same category.

\subsection{How to achieve it ?}
In order to design such visualization dashboard, one need to chose
visualizations to apply to refined datasets in order to achieve some
identified final goal.  This lead to several tasks, independently of
the concrete methodology one chooses to use : \iitem{a} define what
will be the purpose of the dashboard \iitem{b} select data to be
visualize \iitem{c} chose visualization for each group of data
\iitem{d} fill the visualization with the well formatted data set
\iitem{e} arrange the visualisations into a (multiple) dashboard
\iitem{f} audit the dashboard according to the initial motivation

\subsection{How to implement it}
Those tasks refer to three different domains :\\
Requirement Engineering \iitem{a} \& \iitem{f}\\
Data treatment \iitem{b} \& \iitem{d}\\
Dashboard design \iitem{c} \& \iitem{e}\\
The separation of the whole problem in domain specific sub-part aim to
simplify the resolution of each domain challenge and to allow the
contribution of specialist for a domain specific problem without
disturbance from the other part noise.
% existing solution
One could use existing solutions to implement each of those tasks, for
example using SQL would be a suitable choice to store and refine
sensor data, IFML to model the wanted organization of the dashboard
and temporal logic to define the succession of action aimed to be
performed with the final product, and finally D3.JS, AmChart or an
other visualization library to implement the widgets with some HTML5
code to structure the result.
% DSML
One other way to tackle it is to design one DSL per domain This way,
one, when in charge of a given domain, may deal with a language
specifically designed for his task, without being polluted by others
vocabulary or domain constraints.